---
title: "Separate, unite, and strings"
author:
  - [TEAM NAME GOES HERE]
output: 
  html_document:
    df_print: paged
    highlight: textmate 
    theme: readable
    code_folding: hide
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
---

# Overview

These exercises will help you practice applying `separate`, `unite`, and regular expressions. You will use a messy dataset with information about cardiovascular disease (CVD).

## Pre-requisites 

Before starting these exercises, you should have a good understanding of 

1. The [Tidy your data](https://rstudio.cloud/learn/primers/4.1) Primer.

1. Chapter 12.3 - 12.7 and Chapter 14 of [R for Data Science](https://r4ds.had.co.nz/tidy-data.html#pivoting)

## Setup

```{r, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(magrittr)

```

## Data dictionary

```{r}

cvd_messy_descr <-
  c("ID" = 'Participant identification',
    "cvd_status" = "1: CVD event occurred; 0: censored",
    "cvd_time" = "time until CVD event or censoring",
    "sbp" = "Systolic blood pressure, mm Hg",
    "dbp" = "Diastolic blood pressure, mm Hg",
    "bp_meds" = "Using blood pressure medication",
    "age_number" = "Age, years",
    "drink" = "Alcohol use",
    "smoke" = "Cigarette use",
    "hba1c" = "Hemoglobin A1c",
    "diabetes" = "Diabetes status",
    "albuminuria" = "Albuminuria status",
    "bp_midrange" = "SBP 130-139 and DBP <90 or DBP 80-89 and SBP <140",
    "rec_bpmeds_acc_aha" = "BP medication initiation or intensification recommended by the 2017 ACC/AHA BP guidelines",
    "rec_bpmeds_jnc7" = "BP medication initiation or intensification recommended by the JNC7 BP guidelines"
)

# the enframe function transforms a vector into a tibble,
tibble::enframe(cvd_messy_descr) %>% 
  gt::gt(rowname_col = "name") %>%
  gt::tab_stubhead(label = 'Variable name') %>% 
  gt::cols_label(value = 'Variable description') %>% 
  gt::cols_align('left') %>% 
  gt::tab_header(title = 'Description of messy cardiovascular disease data')

```

## Import

```{r}

cvd <- readr::read_rds('data/cvd.rds')
cvd

```

# Problem 1 

Select the variables from `cvd` that are factors, and then apply the `table` function to each variable, separately. Store your results in a list with names corresponding to the variable names. 

```{r}

readr::read_rds('solutions/01_solution.rds')

```

# Problem 2 

Use `bind_rows` to stack the table results on top of each other in a `tibble`. Then, use the `tbl_string()` function to create a string that shows the number of `'Yes'` values and the percent of `'Yes'` values out of the values that are `'Yes'`, `'No'`, or `'Unknown'`. Select just two columns from the resulting data: a column showing variable name and the column showing N(%) values for a table.

__Hints__: 

- the `.id` argument of `bind_rows` will include an additional column in the resulting `tibble` that shows names of the original `list`. 

- `tbl_string()` works just like `glue()`, but automatically rounds numbers for printing in tables. 

```{r}

readr::read_rds('solutions/02_solution.rds')

```

# Problem 3 

Create a `list` with the following __names__ and objects: 

- __Yes__: a subset of `cvd` comprising participants who were using medication to lower blood pressure. 

- __No__: a subset of `cvd` comprising participants who were using medication to lower blood pressure. 

Use the code you wrote in problem 1 and 2 inside of `map` to create a `tibble` similar to the one in problem 2, but with one column per subset in your initial `list`.

__Hints__: I used a `map` inside of a `map` for this.

```{r}

readr::read_rds('solutions/03_solution.rds')

```


# Problem 4

make a simple table to present your results, and then interpret them. Which group of participants is more likely to smoke? Drink? Test positive for diabetes? What may explain these results?

```{r}

readr::read_rds('solutions/04_solution.rds')

```
