---
title: "Iteration"
subtitle: "Introduction to purrr"
author: "Byron C. Jaeger"
date: "Last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/intro_2R_hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, center, middle

```{r, include = FALSE}

library(tidyverse)
library(gt)
library(broom)
library(tblStrings)
library(glue)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 5.5,
  dpi = 72 * 5
)


```

# A tedious task

---

## Multiple means

Suppose I ask you to find the mean value of every numeric variable in the synthetic CVD dataset. 

```{r}

cvd <- read_rds('data/cvd.rds')

mean_status <- mean(cvd$cvd_status, na.rm = TRUE)
mean_time   <- mean(cvd$cvd_time, na.rm = TRUE)
mean_sbp    <- mean(cvd$sbp, na.rm = TRUE)
mean_dbp    <- mean(cvd$dbp, na.rm = TRUE)
mean_age    <- mean(cvd$age_number, na.rm = TRUE)
mean_hba1c  <- mean(cvd$hba1c, na.rm = TRUE)

```

What are the __good__ things about this approach?

- Simple and clear

- Easy to do

- Easy to learn

---

## Multiple means

Suppose I ask you to find the mean value of every numeric variable in the synthetic CVD dataset. 

```{r}

cvd <- read_rds('data/cvd.rds')

mean_status <- mean(cvd$cvd_status, na.rm = TRUE)
mean_time   <- mean(cvd$cvd_time, na.rm = TRUE)
mean_sbp    <- mean(cvd$sbp, na.rm = TRUE)
mean_dbp    <- mean(cvd$dbp, na.rm = TRUE)
mean_age    <- mean(cvd$age_number, na.rm = TRUE)
mean_hba1c  <- mean(cvd$hba1c, na.rm = TRUE)

```

What are the __bad__ things about this approach?

- Repetition increases probability of making a mistake 

- Hard to update and make changes

- Does not scale well

---

## Classic iteration

In the beginning, there was the `for` loop.

- Based on some index, usually denoted as `i`.

- Requires a set of pre-defined index values (i.e., `i` = 1, 2, and 3)

```{r}

for (i in 1:5){
  
  print(i)
  
}

```

---

## Classic iteration

- `for` loop index options are flexible. 

- e.g., loop over a set of character values instead of a set of numbers

```{r}

for (i in names(cvd)){
  
  print(i)
  
}

```

---

## Classic iteration

- `for` loops can make the `mean` task much less tedious.

- loop over all the names and use `if` to act on continuous variables.

```{r}

for (i in names(cvd)) {

  if (is.numeric(cvd[[i]])) {
    print(
      tbl_string('mean of {i}: {mean(cvd[[i]], na.rm = TRUE)}')
    )
  }  
  
}

```

---

## Classic iteration

The hard thing about `for` loops: storing values

```{r}

# initialize empty vectors for results
mean_values <- c()
mean_names <- c()
for (i in names(cvd)) {

  if (is.numeric(cvd[[i]])) {
    # append the new mean to the list
    mean_values <- c(mean_values, mean(cvd[[i]], na.rm = TRUE))
    mean_names <- c(mean_names, i)
  }  
  
}

names(mean_values) <- mean_names

mean_values

```

---

## Tidy iteration

`purrr` is an R package in the tidyverse.

- designed to abstract away some of the extraneous syntax in `for` loops. 

- main function is `map`, which works like a for loop

- works extremely well with `lists` and `dplyr` functions

---

## Tidy iteration

```{r, eval = FALSE}

cvd %>% #<<
  select(where(is.numeric)) %>% 
  map(.f = ~ mean(.x, na.rm = TRUE))

```

```{r, echo = FALSE}
print(cvd, width = 60)
```

---

## Tidy iteration

```{r, eval = FALSE}

cvd %>% 
  select(where(is.numeric)) %>% #<<
  map(.f = ~ mean(.x, na.rm = TRUE))

```

```{r, echo = FALSE}

cvd %>% 
  select(where(is.numeric)) %>% 
  print(width = 60)
```

---

## Tidy iteration

```{r}

cvd %>% 
  select(where(is.numeric)) %>% 
  map(.f = ~ mean(.x, na.rm = TRUE)) #<<

```

---

## Tidy iteration

`map` has variations that can return a specific type of vector, 

- `map_dbl` returns a double vector

```{r}

cvd %>% 
  select(where(is.numeric)) %>% 
  map_dbl(.f = ~ mean(.x, na.rm = TRUE)) #<<

```

---

## Tidy iteration

`map` has variations that can return a specific type of vector, 

- `map_chr` returns a character vector

```{r}

cvd %>% 
  select(where(is.numeric)) %>% 
  map_chr(.f = ~ mean(.x, na.rm = TRUE)) #<<

```

---

## Tidy iteration

`map` has variations that can return a specific type of vector, 

- `map_df` binds results into a tibble.

```{r}

cvd %>% 
  select(where(is.numeric)) %>% 
  map_df(.f = ~ mean(.x, na.rm = TRUE)) #<<

```
---

## Mapping example

Let's say we are trying to guess the regression slope for the following problem. 

```{r}
df <- tibble(x = rnorm(250), y = 3 * x + 1 + rnorm(250, sd = 5))
ggplot(df, aes(x = x, y = y)) + geom_point()

```

---

## Mapping example

We want to select a regression slope value that minimizes the sum of squared differences between predicted values (red line) and observed values (gray points).

```{r}

ggplot(df, aes(x = x, y = y)) +
  geom_point(shape = 21, col = 'black', fill = 'grey') +
  geom_smooth(method = 'lm', col = 'red', se = FALSE)


```

---

## Mapping example

So how about we try a whole bunch of different slope values for our line and see which one minimizes the sum of squared differences?

```{r, echo = FALSE}

ggplot(df, aes(x = x, y = y)) +
  geom_point(shape = 21, col = 'black', fill = 'grey') +
  geom_abline(intercept = 3, col = 'red', slope = 0) + 
  geom_abline(intercept = 3, col = 'red', slope = 1) + 
  geom_abline(intercept = 3, col = 'red', slope = 2) +
  geom_abline(intercept = 3, col = 'red', slope = 3) +
  geom_abline(intercept = 3, col = 'red', slope = 4) +
  geom_abline(intercept = 3, col = 'red', slope = 0.5) + 
  geom_abline(intercept = 3, col = 'red', slope = 1.5) + 
  geom_abline(intercept = 3, col = 'red', slope = 2.5) +
  geom_abline(intercept = 3, col = 'red', slope = 3.5) +
  geom_abline(intercept = 3, col = 'red', slope = 4.5) 

```

---

## Mapping example

first we'll do some single cases

```{r}

estimated_intercept <- 3
estimated_slope <- 1

predictions <- estimated_intercept + estimated_slope * df$x

# mean squared error
sq_error <- (predictions - df$y)^2
mse <- mean(sq_error)
mse

```

---

## Mapping example

first we'll do some single cases

```{r}

estimated_intercept <- 3
estimated_slope <- 1.5

predictions <- estimated_intercept + estimated_slope * df$x

# mean squared error
sq_error <- (predictions - df$y)^2
mse <- mean(sq_error)
mse

```

---

## Mapping example

Now we `map`.

```{r}

slopes <- seq(1, 5, length.out = 1000)

results <- map_dbl(
  .x = slopes,
  .f = ~ {
    estimated_intercept <- 3
    estimated_slope <- .x
    
    predictions <- estimated_intercept + estimated_slope * df$x
    
    # mean squared error
    sq_error <- (predictions - df$y)^2
    mse <- mean(sq_error)
    mse
  }
)



```

---

## Mapping example

and then `ggplot`.


```{r, echo = FALSE}

library(ggforce)
tibble(
  slp_estm = slopes,
  mse_vals = results
) %>% 
  ggplot(
    aes(
      x = slp_estm, 
      y = mse_vals
    )
  ) + 
  geom_point(shape = 21, fill = 'red', col = 'grey') + 
  geom_mark_circle(
    aes(
      filter = mse_vals == min(mse_vals),
      label = paste('Slope estimate =', round(slp_estm,1)),
      description = paste('conditional on knowing the intercept')
    ),
  ) + 
  labs(y = 'Mean squared error', x = 'Estimated slope')

```

---

## Learning more

To learn more, see

- The excellent purrr [tutorials](https://jennybc.github.io/purrr-tutorial/)

- The purrr [website](https://purrr.tidyverse.org/index.html)


